---
title: "Quick start guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

corella is a tool for standardising data in R to use the [*Darwin Core standard*](https://www.gbif.org/darwin-core). Darwin Core standard is the primary data standard for species occurrence data---records of what organisms were observered in a location and time---in the Atlas of Living Australia, other Living Atlases and GBIF. The standard allows the ability to compile data from a variety of sources, improving the ease to share, use and reuse data.

The main tasks to standardise data to comply with Darwin Core standards is to 

  1. Ensure columns use valid Darwin Core terms as column names
  2. Include all required information (e.g. scientific name, unique observation ID, valid date)
  3. Ensure columns contain valid data

If this sounds daunting, corella is designed to reduce confusion of how to get started and which Darwin Core terms might match your column names.

# Install

To install from CRAN:

```{r}
#| eval: false
install.packages("corella")
```

To install the development version from GitHub:

```{r}
#| eval: false
#| message: false
#| warning: false
# install.packages("devtools")
devtools::install_github("AtlasOfLivingAustralia/corella")
```

To load the package:

```{r}
library(corella)
```


# Rename, add or edit columns

Here is a minimal example dataset of [some species] observations. In our dataframe `df` there are columns that contain information that we would like standardised using Darwin Core.

```{r}
#| code-fold: true
#| 
library(tibble)
library(lubridate)

df <- tibble(
  latitude = c(-35.310, "-35.273"), # deliberate error for demonstration purposes
  longitude = c(149.125, 149.133),
  date = c("14-01-2023", "15-01-2023"),
  time = c("10:23:00", "11:25:00"),
  month = c("January", "February"),
  day = c(100, 101),
  species = c("Callocephalon fimbriatum", "Eolophus roseicapilla"),
  n = c(2, 3),
  crs = c("WGS84", "WGS8d"),
  country = c("Australia", "Denmark"),
  continent = c("Oceania", "Europe")
  )
```


```{r}
df
```

We can standardise our data with `use_` functions. These functions a prefix of `use_`, and a suffix named to suggest what type of data they are used to standardise. For example, `use_coordinates` or `use_datetime`. The argument names in `use_` functions are valid Darwin Core terms. Our hope is that by grouping Darwin Core terms into groups based on their data type, corella makes it easier for users to know which Darwin Core terms to use as column names (which is one of the most confusing parts of Darwin Core).

Let's specify that the scientific name (i.e. genus & species name) in our data is in the `species` column by using `use_scientific_name()` to update our dataframe.

```{r}
df |>
  use_scientific_name(scientificName = species)
```

You'll notice 2 things happen:

  1. The `species` column in our dataframe has been renamed to `scientificName`.
  2. `use_scientific_name()` ran a check on our `species` column to make sure it was formatted correctly.

What happens when we add a column with an error in it? In our `df`, the `latitude` column is "accidentally" a class `character` column, when it should be a `numeric` column. When we try to add it using `use_coordinates()`, an error will tell us something is wrong.

```{r}
#| eval: true
#| error: true
# is it possible to print this error?
df |>
  use_scientific_name(scientificName = species) |>
  use_coordinates(decimalLongitude = longitude,
                  decimalLatitude = latitude)
```


## Fix or update columns

How do we fix the column that caused an error? One way is to edit the column within the `use_` function containing our problematic column.

Each `use_` function is essentially a specialised [`mutate()`]() function from [dplyr](). This means you can edit columns using the same process you would use with `mutate()`. Let's fix the `latitude` column so that it is class `numeric`, which will fix the error.

```{r}
df |>
  use_scientific_name(scientificName = species) |>
  use_coordinates(decimalLongitude = longitude,
                  decimalLatitude = as.numeric(latitude))
```

## Auto-detect columns

corella is also able to detect when columns already contain valid Darwin Core terms as column names. For example, `df` contains columns with locality information. We can add `use_locality()` to our pipe to specify and check these columns. Some of our columns already have valid Darwin Core terms as column names (`country` and `continent`). If we add `use_locality()` to our pipe without any arguments, it will detect these valid Darwin Core columns in `df` and check them automatically.

```{r}
df |>
  use_scientific_name(scientificName = species) |>
  use_coordinates(decimalLongitude = longitude,
                  decimalLatitude = as.numeric(latitude)) |>
  use_locality()
```

By auto-detecting columns, corella prevents users from having to specify every single column when they are already named correctly in their dataset. This functionality will definitely help reduce the workload on datasets with lots of columns!


# Suggest a workflow

Unsure where to start? Confused about the minimum requirements to share your data? Try using `suggest_workflow()`.

`suggest_workflow()` provides a high level summary designed to show:

  1. Which column names match valid Darwin Core terms
  2. The minimum requirements for data to be shared, and which columns are present in your dataframe.
  3. A suggested workflow to help you add the minimum required columns
  4. Additional functions you could add to your piped workflow (based on column names in your dataset that match other Darwin Core terms)
  
The intention of `suggest_workflow()` is to provide a general help function whenever users feel uncertain about what to do next. Let's see what it says about what to do with our original dataframe `df`.

```{r}
df |>
  suggest_workflow()
```

`suggest_workflow()` will update the suggested function pipe to only suggest functions that are necessary to standardise your data correctly. 

For example, after using one of the suggested functions `use_occurrences()`, if we run `suggest_workflow()` again, the output message no longer suggests `use_occurrences()`. 

```{r}
#| message: false
#| warning: false
df_edited <- df |>
  use_occurrences(
    occurrenceID = random_id(),
    basisOfRecord = "humanObservation"
    )
```

```{r}
df_edited |>
  suggest_workflow()
```

# Test your data

If your dataset already uses valid Darwin Core terms as column names, instead of working through each `use_` function, you might wish to run tests on your entire dataset. To run checks on your data like a test suite, use `check_dataset()`. Much like `devtools::test()` or `devtools::check()`, `check_dataset()` runs the relevant check on each matching Darwin Core column and returns a summary of the results, along with any error messages returned by those checks.

```{r}
df <- tibble(
  latitude = c(-35.310, "-35.273"), # deliberate error for demonstration purposes
  longitude = c(149.125, 149.133),
  date = c("14-01-2023", "15-01-2023"),
  individualCount = c(0, 2),
  species = c("Callocephalon fimbriatum", "Eolophus roseicapilla"),
  country = c("AU", "AU"),
  occurrenceStatus = c("present", "present")
  )

df |>
  check_dataset()
```


# `use_measurements`: a work in progress

Let's use a small sample of a real dataset of Australian native plant species, including Eucalypts and Acacias. The dataset contains lots of columns containing measurements that capture the traits or environment of each species occurrence. I have extracted 3 examples of these measurement columns: `LMA_g.m2`, `LeafN_area_g.m2`, `PNUE`.

```{r}
#| warning: false
#| message: false
library(tibble)
library(readr)
library(dplyr)
library(tidyr)
library(here)

# take a small sample
df_filtered <- read_csv(here("inst", "examples", "data", "westerband_2022_wdate.csv")) |>
  select(Site, Species, Latitude, Longitude, LMA_g.m2, LeafN_area_g.m2, PNUE)

df_filtered
```


The way that Darwin Core handles measurement fields is slightly different to how they are organised in these data. In `df_filtered`, lots of information about what data the column contains is in the abbreviated column name. This includes the type of measure, the unit it is measured in, and the value itself. In Darwin Core, this information needs to be documented individually for each measurement, which requires multiple columns and multiple rows (ie a 'long' data format). 

corella can convert measurement columns into Darwin Core standard with `use_measurements()`. This function requires the column, unit and measure type are specified for each respective measurement, after which `use_measurements()` assigns this information to the correct Darwin Core columns. `use_measurements()` then nests them so that they information is linked to each species occurrence, held in a nested column called `measurementOrFact`.

```{r, message=FALSE, warning=FALSE}
df_nested <- df_filtered |>
  slice(220:270) |>
  use_measurements(cols = c(LMA_g.m2, 
                            LeafN_area_g.m2, 
                            PNUE),
                   unit = c("g/m2", 
                            "g/m2", 
                            "co2/s"),
                   type = c("leaf mass per area", 
                            "leaf nitrogen per area", 
                            "photosynthetic nitrogen use efficiency")
                   )


df_nested |> slice(1:3)
```


To view or access this data, you can use `dplyr::unnest()`, which will extract the information into a 'long' data format.

```{r}
df_nested |>
  unnest(measurementOrFact)
```

